Lab 2 Final Report: COVID-19 Case Rate vs Population Demographics and Mask Policy

### 2020-12-05 Jackson_Liu_Temlock_Zhang_Lab2_Draft_Report.Rmd
Updated as discussed and knitted a PDF

## Assumptions

@@

## Model Variables

@@

## Model 1

@@

### Model 1 Regression

@@

### Model 1 Limitations

@@

The final CLM assumption examined for model 1 is the assumption of normality of errors. This is a necessary assumption to ascertain that the errors used in the regression are drawn from a normal distribution, so that they can be accepted when used to calculate the significance levels. Both a QQ plot and histogram of the distribution of the residuals is plotted to observe the normality.

```{r, message = FALSE}
# Create a QQ plot for the model 1 residuals
plot(model1, which=2)

# Create a histogram of the distribution of the model 1 residuals
df %>%
  ggplot(aes(x = resid(model1))) +
  stat_bin() +
  geom_histogram(fill = 'skyblue', color = 'grey30', bins=30) +
  labs(title =  "Histogram of the Model 1 Residuals", 
       x = "Model 1 residuals", y = 'Count')
```

@@

## Model 2

@@

### Model 2 Regression

@@

### Model 2 Limitations

#### i. IID Sampling 
(This will be discussed in the following section - 3. Limitations of the Models)

#### ii.Linear Conditional Expectation
To assess the linear conditional expectation assumption in higher-dimensional space, the plots of predicted versus residuals of the model 2 is investigated as follows.

@@

#### iii.No Perfect Collinearity

@@

#### iv.Homoskedastic Errors
To evaluate the homoskedastic errors assumption, the square root of the residuals is plotted against the fitted values of model 2 (scale-location plot). Although, the curve is not perfectly flat, there is no obvious variance of the errors either. For reference, the data points marked in the plot represent Florida (10), New York (33) and North Dakota (35).

@@

plot_one / plot_two
```
Limited by the sample size, it is hard to observe a perfect normal distribution from the histogram. However, the qqplot implies no deviation from normality. Overall, the normally distributed errors assumption is satisfied for model 2.

### Cook Distance

Additionally, cook distance is used to estimate the influence of each single data point.

```{r, echo =FALSE}
plot(model2, which = 5)
```
No obvious outlier could be detected here. For reference, the marked data points are Florida(10), Maine(20), and North Dakota(30).

## Model 3

@@

### Model 3 Regression

@@

### CLM Assumptions Validation for Model 3:

#### I.I.D Assumption

( Discussed in the common section)

#### Zero Conditional Mean

@@

#### Normality of Errors

From the Normal Q-Q plot below, the points generally stay close to the normal line, with the tails exhibit certain degree of skew. There is no significant evidence of concerning normality in the errros.
```{r}
plot(model3, which=2)
```

#### Homoskedasticity

@@

#### Collinearity

#### Cook's Distance

From the Cook's plot, there is no strong evidence of concerning outliers.

@@

## 3. Limitations of the Models

The most general requirement for all the linear models is that data points be independent and identically distributed (IID) which they are built off of. While for the most part the events of one state will primarily affect that single state, travel by individuals spreading the virus across state lines is an inevitable occurence. This may cause the spread of a virus within one state to be influenced by those around it, or generally by its position in the country. In addition, all states are involved in the same market economy system of the country to varying degrees, even though there are economic performance divisions across the states. Likewise, certain regions of the country may also have similar demographics due to a shared history across general geographic regions. While this is acknowledged to occur, the effect is likely minor compared to the policy data collected state by state which would affect the spread of the virus. Therefore it is relatively safe for this assumption to be met.

In examining the results of various diagnostics across different models, generally the CLM assumptions are best met with the second model. In the first model, deviations away from certains assumptions are the strongest. This is seen mostly with regards to the linear conditonal expectation assumption, where the residuals against the fitted values have a noticeable inverted "V" shape as opposed to being constant around zero. In the second model, the same diagnostic shows that the assumption is better met.

@@

## 4. Regression Table

@@

Initially addressing the progression of the models at a macro level, the regression table highlights the fact that as the models build upon one another, they seem to get better at explaining the relationship between the input and output variables. Between model 1, 2, and 3, both the R2 and stricter adjusted R2 value increases as variables are added. The R2 value reflects the amount of variation that can be explained in the model on a 0-1 scale, while the adjusted R2 value does the same but also takes into account the number of variables measured (as can be seen in the decreasing degrees of freedom, df). The adjusted R2 value jumps from 0.082 initially in model 1 to a value of 0.267 in model 2, which is more than a 200% increase. The value again increases, albeit by a smaller amount, between model 2 and a value of 0.376 in model 3, which represents a 40% jump. When the R2 value is not adjusted, this value increases to 0.439 for model 3. Although not extraordinarily high given the scale, the value shows that that model 3 still explains much of the variance and is a relatively good predictor of the data, especially when compared to model 1. An additional value that reflects this improvement is the fact that the F-statistic is significant for all three models. The F-statistic is used to test whether or not to use the full model (inclusive of all input variables) or a reduced one with fewer variables. Given that the null hypothesis for this test is that the reduced model should be over the full model, this can be safely rejected in favor of the full model.

@@

In model 1, the coefficient of -0.174 for the senior rate per 100,000 is significant with a p-value of 0.0234 (per the regression results from the Model 1 section). When the covariate variables of unemployment rate and poverty rate are added in to model 2, the coefficient for senior rate adjusts to a greater magnitude of -0.231 while the p-value appreciably decreases to a value of 0.0015 (per the regression results from the Model 2 section). This explains that not only has the magnitude of effect of the senior rate on case rate increased, but that this increased effect is also more statistically significant. In addition, the coefficients of the covariate variables themselves are also highly significant with p-values of 0.0005 and 0.0059 for poverty rate and unemployment rate respectively. The magnitude of the effect of poverty rate is almost equivalent to that of the senior rate, with a single increase in poverty rate per 100,000 equivalent to 0.232 per 100,000 for case rate, while the unemployment rate has stronger negative coefficient of -0.486 per 100,000 (roughly twice the magnitude of the other two variables).

Moving onto model 3, the magnitude of the coefficient for senior rate increases once again to a value of -0.242 while the statistical significance increases with a decreased p-value of 0.0004 (per the regression results from the Model 3 section). As highlighted before, the more complex model does a better job of predicting the effect of the senior rate variable. Conversely, the p-values for the poverty rate and unemployment rate coefficients become less significant, as they increase to 0.002 and 0.049 respectively. There are also detectable changes in the coefficient values, as the magnitude for both decreases to 0.197 and -0.364 respectively. The changes in the p-values are likely a direct result of the added variables of mask policy and the logarithmic transform of population density, where these variables introduce some additional variance in the model 2 covariates. Notably, the coefficient for the mask policy covariate is significant with a p-value of 0.0027 and has by far the greatest effect on the outcome variable with a coefficient of -945.6. The populuation density covariate is not significant with a p-value of 0.3369.

In terms of practical significance, there is great discrepancy between the effect of the mask policy variable compared to the other three significant features. With a coefficient of -945.6, this means that given the binary mask policy variable, when a mask policy is implemented (value of 1), there is an decrease in case rate of roughly 945 per 100,000 compared to when the policy is not in place. The next largest effect is from the unemployment rate variable, where in the final model, the coefficient of -0.364 translates as a decrease of 0.364 cases per 100,000 for every additional increase of 1 in unemployment rate per 100,00. More intuitively, it can be said that there is a decrease of roughly 1 case per 100,000 with every 3 additional people unemployed per 100,000. This effect is roughly 2,600 times smaller than the effect of implementing a mask policy. Similarly, the effects of both poverty rate and senior rate are nominal. For every 5 additional people per 100,000 who are below the poverty line, there is roughly 1 additional case per 100,000, and for every 4 additional seniors per 100,000 within the state population, there is roughly 1 fewer case per 100,000. In comparison with likewise relatively minimal effect of the unemployment rate, these absolute effects on case rate are approximately one-third (1/3) and one-half (1/2) smaller respectively.

## 5. Omitted Variables

@@

### 5-1: Urban vs Rural Population Ratio

One of the variables that was hypothesized as being potentially influential is a ratio of a state's urban to rural population. Currently, the dataset contains population information as averaged across each state as a whole. While population density is potentially useful, it was found to not be significant in the models. This may be due to states which have large urban centers, where population density woukd be high, losing their impact due to also containing large areas of land that is sparsely populated. One example of this would be New York, which contains several of the most densely populated counties in the country, but also large rural areas.

@@

### 5-2: Median Income

@@

### 5-3: Gender Ratio

The ratio of gender between women and men may impact the COVID case rate in that men are more suscepitble to viral infections and generally have lower life expectancies from lifestyle differences. Greater susceptibility may relate to the case rate in that men may more often become infected when exposed to COVID under similar circumstances as a women. Their shorter life expentancy may also impact the case rate in that more COVID cases may result in greater or more serious symptoms, with less occurances of an infection that is asymptomatic or goes undetected. In either of these cases, a higher fraction of men among the population would lead to an overall higher COVID case rate.

@@

### 5-4: Mask Use Among Population

@@

### 5-5: GDP Percentage of Tourism

@@

## 6. Conclusion

@@

Starting with the variable with the greatest effect size, it can be seen that the implementation of a mask policy in a state leads to a decrease of 945 cases per 100,000. Given that research from leading health organizations shows that the wearing of masks helps to decrease the chance of the transmission of COVID-19, it could be suggested that a mask policy is effective in increasing the adoption of wearing masks among the state population. Next, it can be seen that there is a decrease of roughly 1 case per 100,000 with every 3 additional people unemployed per 100,000. This suggests that as unemployment increases, the case rate decreases, which could be a result of the fact that those who are unemployed may be less mobile as they are not commuting into work, lowering the likelihood of contracting the virus. Additionally, at the time that this data was collected, the Coronavirus Aid, Relief, and Economic Security (CARES) Act was still supporting the unemployed with relief payments so there was less urgency to be out looking for work. Following unemployment, when examining the poverty variable there is actually an increase of roughly 1 case per 100,000 for every 5 additional people per 100,000 living under the federal poverty line. Although there is some intuition behind the unemployment and poverty rates having similar effects on case rate, it can be seen from the test of collinearity in the Model 2 section that these two variables are not explaining the same effects. This difference is likely due to the fact that those living below the poverty line have less access to suitable living conditions and sanitary goods that help to alleviate the possibility of COVID-19 infection, while those who were recently employed up until the pandemic hit may not be in the same economical situation. Finally, the effect of the senior rate is negative, with an additional 4 seniors per 100,000 translating to roughly 1 fewer case per 100,000. As touched upon in the introduction, this could be largely down to the fact that it is widely broadcasted that seniors fall within the high risk category in terms of having severe complications from COVID-19, and therefore they are more diligent in the precautions they take against contracting the virus.

In conclusion, we see that although the population demographics had varying effects on the COVID-19 case rate, the greatest effect was seen from the mask policy. Going back to our original question, it can be suggested within the data that the effect of a mask-related policy is much larger than that of population demographics. Therefore, it seems that the COVID-19 case rate at the state-level is much more dependent on masks, and that ultimately, there are effective policy and control measures that can be taken by state governments that have a greater contribution towards combating it's spread relative to factors that cannot be controlled, such as population demographics. However, it should be noted that these insights are qualified by the assumptions and limitations laid out throughout the report, namely that there are potential issues in meeting the IID assumption and that there is explanatory information that is lost from both the variables investigated as well as the those that were not included in the dataset.

>>>>

## 1-1. Assumptions

@@

@@

## 2-1. Model 1

### 2-1-1. Model 1 Exploratory Data Analysis

@@

### 2-1-2. Model 1 Regression

@@

### 2-1-3. Model 1 Limitations

@@

The final CLM assumption examined for model 1 is the assumption of normality of errors. This is a necessary assumption to ascertain that the errors used in the regression are drawn from a normal distribution, so that they can be accepted when used to calculate the significance levels. Both a Q-Q plot and histogram of the distribution of the residuals is plotted to observe the normality.

```{r, message = FALSE}
# Create a histogram of the distribution of the model 1 residuals
plot_one <- df %>%
  ggplot(aes(x = resid(model1))) +
  stat_bin() +
  geom_histogram(fill = 'skyblue', color = 'grey30', bins=30) +
  labs(title =  "Histogram of the Model 1 Residuals", 
       x = "Model 1 residuals", y = 'Count')

# Create a QQ plot for the model 1 residuals
plot_two <- df %>% 
  ggplot(aes(sample = resid(model1))) + 
  stat_qq() + stat_qq_line()

plot_one / plot_two
```

@@

## 2-2. Model 2

### 2-2-1. Model 2 Exploratory Data Analysis

@@

### 2-2-2. Model 2 Regression

@@

### 2-2-3. Model 2 Limitations

#### ii. Linear Conditional Expectation
To assess the linear conditional expectation assumption in higher-dimensional space, the plots of predicted versus residuals of the model 2 is investigated as follows.

@@

@@

#### iv. Homoskedastic Errors
To evaluate the homoskedastic errors assumption, the square root of the residuals is plotted against the fitted values of model 2 (scale-location plot). Although, the curve is not perfectly flat, there is no obvious variance of the errors either. For reference, the data points marked in the plot represent Florida (10), New York (33) and North Dakota (35).

@@

plot_one / plot_two
```
Limited by the sample size, it is hard to observe a perfect normal distribution from the histogram. However, the Q-Q plot implies no deviation from normality. Overall, the normally distributed errors assumption is satisfied for model 2.

Additionally, Cook's distance is used to estimate the influence of each single data point.

```{r, echo =FALSE}
plot(model2, which = 5)
```

No obvious outliers could be detected here. For reference, the marked data points are Florida (10), Maine (20), and North Dakota (30).

## 2-3. Model 3

### 2-3-1. Model 3 Exploratory Data Analysis

@@

### 2-3-2. Model 3 Regression

@@

### 2-3-3. Model 3 Limitations

#### i. I.I.D Assumption

@@

@@

To access the assumption of normality of error distribution, both the histogram and the Q-Q plot of the residuals is shown as follow.

```{r, echo=FALSE, warning=FALSE, message=FALSE}
plot_one <- df %>% 
  ggplot(aes(x = resid(model3))) +
  geom_histogram(fill = 'skyblue', color = 'grey30', bins=30) +
  labs(title =  "Histogram of the Model 3 Residuals", 
       x = "Model 3 residuals", y = 'Count')

plot_two <- df %>% 
  ggplot(aes(sample = resid(model3))) + 
  stat_qq() + stat_qq_line()

plot_one / plot_two
```

There is a fairly normal distribution that can be seen from the histogram above. From the Q-Q plot, the points generally stay close to the normal line, with the tails exhibiting a certain degree of skew. However, in general, it implies no deviation from normality. Overall, the normally distributed errors assumption is satisfied for model 3.

From the Cook's distance plot, there is no strong evidence of concerning outliers. For reference, the marked data points are D.C. (9), New Hampshire (30), and North Dakota (35).

@@

# 3. Limitations of the Models

The most general requirement for all the linear models is that data points be independent and identically distributed (IID) which they are built off of. While for the most part the events of one state will primarily affect that single state, travel by individuals spreading the virus across state lines is an inevitable occurrence. This may cause the spread of a virus within one state to be influenced by those around it, or generally by its position in the country. In addition, all states are involved in the same market economy system of the country to varying degrees, even though there are economic performance divisions across the states. Likewise, certain regions of the country may also have similar demographics due to a shared history across general geographic regions. While this is acknowledged to occur, the effect is likely minor compared to the policy data collected state by state which would affect the spread of the virus. Therefore it is relatively safe for this assumption to be met.

In examining the results of various diagnostics across different models, generally the CLM assumptions are best met with the second model. In the first model, deviations away from certains assumptions are the strongest. This is seen mostly with regards to the linear conditional expectation assumption, where the residuals against the fitted values have a noticeable inverted "V" shape as opposed to being constant around zero. In the second model, the same diagnostic shows that the assumption is better met.

@@

# 4. Regression Table

@@

Initially addressing the progression of the models at a macro level, the regression table highlights the fact that as the models build upon one another, they seem to get better at explaining the relationship between the input and output variables. Between model 1, 2, and 3, both the R2 and stricter adjusted R2 value increases as variables are added. The R2 value reflects the amount of variation that can be explained in the model on a 0-1 scale, while the adjusted R2 value does the same but also takes into account the number of variables measured (as can be seen in the decreasing degrees of freedom, df). The adjusted R2 value jumps from 0.082 initially in model 1 to a value of 0.267 in model 2, which is more than a 200% increase. The value again increases, albeit by a smaller amount, between model 2 and a value of 0.376 in model 3, which represents a 40% jump. When the R2 value is not adjusted, this value increases to 0.439 for model 3. Although not extraordinarily high given the scale, the value shows that that model 3 still explains much of the variance and is a relatively good predictor of the data, especially when compared to model 1. An additional value that reflects this improvement is the fact that the F-statistic is significant for all three models. The F-statistic is used to test whether or not to use the full model (inclusive of all input variables) or a reduced one with fewer variables. Given that the null hypothesis for this test is that the reduced model should be over the full model, this can be safely rejected in favor of the full model. Additionally, it can be seen that the standard errors for the significant variables are relatively small and are consistent across the models.

@@

In model 1, the coefficient of -0.174 for the senior rate is significant with a p-value of 0.0234 (per the regression results from the Model 1 section). When the covariate variables of unemployment rate and poverty rate are added in to model 2, the coefficient for senior rate adjusts to a greater magnitude of -0.231 while the p-value appreciably decreases to a value of 0.0015 (per the regression results from the Model 2 section). This explains that not only has the magnitude of effect of the senior rate on case rate increased, but that this increased effect is also more statistically significant. In addition, the coefficients of the covariate variables themselves are also highly significant with p-values of 0.0005 and 0.0059 for poverty rate and unemployment rate respectively. The magnitude of the effect of poverty rate is almost equivalent to that of the senior rate, with a coefficient of 0.232, while the unemployment rate has stronger negative coefficient of -0.486 (roughly twice the magnitude of the other two variables).

Moving onto model 3, the magnitude of the coefficient for senior rate increases once again to a value of -0.242 while the statistical significance increases with a decreased p-value of 0.0004 (per the regression results from the Model 3 section). As highlighted before, the more complex model does a better job of predicting the effect of the senior rate variable. Conversely, the p-values for the poverty rate and unemployment rate coefficients become less significant, as they increase to 0.002 and 0.049 respectively. There are also detectable changes in the coefficient values, as the magnitude for both decreases to 0.197 and -0.364 respectively. The changes in the p-values are likely a direct result of the added variables of mask policy and the logarithmic transform of population density, where these variables introduce some additional variance in the model 2 covariates. Notably, the coefficient for the mask policy covariate is significant with a p-value of 0.0027 and has by far the greatest effect on the outcome variable with a coefficient of -945.6. The population density covariate is not significant with a p-value of 0.3369.

In terms of practical significance, there is great discrepancy between the effect of the mask policy variable compared to the other three significant features. With a coefficient of -945.6, this means that given the binary mask policy variable, when a mask policy is implemented (value of 1), there is an decrease in case rate of roughly 945 per 100,000 compared to when the policy is not in place. The next largest effect is from the unemployment rate variable, where in the final model, the coefficient of -0.364 translates to a decrease of roughly 1 case per 100,000 with every 3 additional people unemployed per 100,000. Similarly, the effects of both poverty rate and senior rate are nominal. For every 5 additional people per 100,000 who are below the poverty line, there is roughly 1 additional case per 100,000, and for every 4 additional seniors per 100,000 within the state population, there is roughly 1 fewer case per 100,000.

# 5. Omitted Variables

@@

## 5-1. Urban vs Rural Population Ratio

One of the variables that was hypothesized as being potentially influential is a ratio of a state's urban to rural population. Currently, the dataset contains population information as averaged across each state as a whole. While population density is potentially useful, it was found to not be significant in the models. This may be due to states which have large urban centers, where population density would be high, losing their impact due to also containing large areas of land that is sparsely populated. One example of this would be New York, which contains several of the most densely populated counties in the country, but also large rural areas.

@@

## 5-2. Median Income

@@

## 5-3. Gender Ratio

The ratio of gender between women and men may impact the COVID case rate in that men are more susceptible to viral infections and generally have lower life expectancy from lifestyle differences. Greater susceptibility may relate to the case rate in that men may more often become infected when exposed to COVID under similar circumstances as a women. Their shorter life expectancy may also impact the case rate in that more COVID cases may result in greater or more serious symptoms, with less occurrences of an infection that is asymptomatic or goes undetected. In either of these cases, a higher fraction of men among the population would lead to an overall higher COVID case rate.

@@

## 5-4. Mask Use Among Population

@@

## 5-5. GDP Percentage of Tourism

@@

# 6. Conclusion

@@

Starting with the variable with the greatest effect size, it can be seen that the implementation of a mask policy in a state leads to a decrease of 945 cases per 100,000. Given that research from leading health organizations shows that the wearing of masks helps to decrease the chance of the transmission of COVID-19, it could be suggested that a mask policy is effective in increasing the adoption of wearing masks among the state population. Next, it can be seen that there is a decrease of roughly 1 case per 100,000 with every 3 additional people unemployed per 100,000. This suggests that as unemployment increases, the case rate decreases, which could be a result of the fact that those who are unemployed may be less mobile as they are not commuting into work, lowering the likelihood of contracting the virus. Additionally, at the time that this data was collected, the Coronavirus Aid, Relief, and Economic Security (CARES) Act was still supporting the unemployed with relief payments so there was less urgency to be out looking for work. Following unemployment, when examining the poverty variable there is actually an increase of roughly 1 case per 100,000 for every 5 additional people per 100,000 living under the federal poverty line. Although there is some intuition behind the unemployment and poverty rates having similar effects on case rate, it can be seen from the test of collinearity in the Model 2 section that these two variables are not explaining the same effects. This difference is likely due to the fact that those living below the poverty line have less access to suitable living conditions and sanitary goods that help to alleviate the possibility of COVID-19 infection. As a caveat, it should be noted that these the poverty and unemployment figures are from a 2018 survey, so it does not reflect the full picture of the state demographics at the time of when the COVID-19 related data was captured. Finally, the effect of the senior rate is negative, with an additional 4 seniors per 100,000 translating to roughly 1 fewer case per 100,000. As touched upon in the introduction, this could be largely down to the fact that it is widely broadcasted that seniors fall within the high risk category in terms of having severe complications from COVID-19, and therefore they are more diligent in the precautions they take against contracting the virus.

In conclusion, we see that although the population demographics had varying effects on the COVID-19 case rate, the greatest effect was seen from the mask policy. Going back to our original question, it can be suggested within the data that the effect of a mask-related policy is something that should not be neglected in comparison to the population demographics. Therefore, it seems that the COVID-19 case rate at the state-level is dependent on masks, and that ultimately, there are effective policy and control measures that can be taken by state governments that have some contribution towards combating it's spread relative to factors that cannot be controlled. However, it should be noted that these insights are qualified by the assumptions and limitations laid out throughout the report, namely that there are potential issues in meeting the IID assumption and that there is explanatory information that is lost from both the variables investigated as well as the those that were not included in the dataset.

# References
